A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data

### 2024-09-18 manuscript/manuscript.Rmd
update manuscript

---
title             : "A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data"
shorttitle        : "Tutorial on hazard analysis"
author: 
  - name          : "Sven Panis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "ETH GLC, room G16.2, Gloriastrasse 37/39, 8006 Zürich"
    email         : "sven.panis@hest.ethz.ch"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Richard Ramsey"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "ETH Zürich"
authornote: |
  Neural Control of Movement lab, Department of Health Sciences and Technology (D-HEST).
  <!-- Enter author note here. -->
abstract: |
  Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. To move beyond mean performance measures, various distributional analyses have been proposed. Here we focus on one particular distributional analysis known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. Across four tutorials that we make publicly available on Github and OSF, we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. Along the way, we discuss how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time survival analysis.

@@

```{r setup, include = FALSE}
# Tutorial 1 and 4
pkg1 <- c("papaja", "citr", "tidyverse", "RColorBrewer", "patchwork")

@@

# Tutorial 3

@@

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating mean average performance across a series of trials. However, differences in means conceal when an experimental effect starts, how long it lasts, and whether its onset is time-locked to other events. Such information is useful not only for interpretation, but also for cognitive psychophysiology and computational model selection [@panisAnalyzingResponseTimes2020].
In this tutorial we focus on a distributional method for analyzing time-to-event data that is known as discrete-time event history analysis (EHA), a.k.a. survival, hazard, duration, failure-time, and transition analysis [@singerAppliedLongitudinalData2003; @allisonDiscreteTimeMethodsAnalysis1982; @allisonSurvivalAnalysisUsing2010]. Across four tutorials that we make publicly available on Github and the Open Science Framework (OSF), we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4.

To apply EHA, one must be able to define the event of interest (any qualitative change that can be situated in time, e.g., a button press, saccade onset, fixation offset), time point zero (e.g., target stimulus onset, fixation onset), and measure the passage of time between time point zero and event occurrence in discrete or continuous time units.

@@

Similarly, after dividing time in discrete, contiguous time bins indexed by t, let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular experimental condition. Discrete-time EHA focuses on the discrete-time hazard function h(t) = P(RT = t| RT $\geq$ t) and the discrete-time survivor function S(t) = P(RT $>$ t) = [1-h(t)].[1-h(t-1)].[1-h(t-2)]...[1-h(1)], and not on the probability mass function p(t) = h(t).S(t-1) and the cumulative distribution function F(t) = 1-S(t). The discrete-time hazard probability function gives you the probability that the event occurs (sometime) in bin t, given that the event has not occurred yet in previous bins. Unlike the hazard function, which assesses the unique risk associated with each time bin, the survivor function cumulates the bin-by-bin risks of event *non*occurrence.
For two-choice RT data, the discrete-time hazard function can be extended with the conditional accuracy function ca(t) = P(correct | RT = t), which gives you the probability that a response is correct given that it has been emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. This latter function is also known as the micro-level speed-accuracy tradeoff function.

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons. First, as discussed by @holdenDispersionResponseTimes2009, “probability density functions can appear nearly identical, both statistically and to the naked eye, and yet are clearly different on the basis of their hazard functions (but not vice versa). Hazard functions are thus more diagnostic than density functions” (p. 331).
Second, because RT distributions may differ from one another in multiple ways, @townsendTruthConsequencesOrdinal1990 developed a dominance hierarchy of statistical differences between two arbitrary distributions A and B. For example, if F~A~(t) > F~B~(t) for all t, then both cumulative distribution functions are said to show a complete ordering. Townsend (1990) showed that a complete ordering on the hazard functions —$\lambda$~A~(t) > $\lambda$~B~(t) for all t— implies a complete ordering on both the cumulative distribution and survivor functions —F~A~(t) > F~B~(t) and S~A~(t) < S~B~(t)— which in turn implies an ordering on the mean latencies —mean A < mean B. In contrast, an ordering on two means does not imply a complete ordering on the corresponding F(t) and S(t) functions, and a complete ordering on these latter functions does not imply a complete ordering on the corresponding hazard functions. This means that stronger conclusions can be drawn from data when comparing the hazard functions using EHA. For example, when mean A < mean B, the hazard functions might show a complete ordering (i.e., for all t), a partial ordering (e.g., only for t > 300 ms, or only for t < 500 ms), or they may cross each other one or more times.
Third, EHA does not discard right-censored observations when estimating hazard functions, that is, trials for which we do not observe a response during the data collection period so that we only know that the RT must be larger than some value. This is important because although a few right-censored observations are inevitable in most RT tasks, a lot of right-censored observations are expected in experiments on masking, the attentional blink, and so forth.
Fourth, hazard modeling allows incorporating time-varying explanatory covariates such as heart rate, electroencephalogram (EEG) signal amplitude, gaze location, etc. (Allison, 2010) which is useful for cognitive psychophysiology [@meyerModernMentalChronometry1988].
Finally, as explained by @kelsoOutlineGeneralTheory2013, it is crucial to first have a precise description of the macroscopic behavior of a system (here: h(t) and ca(t) functions) in order to know what to derive on the microscopic level. For example, fitting parametric functions or computational models to data without studying the shape of the h(t) and ca(t) functions can miss important features in the data [@panisStudyingDynamicsVisual2020; @panisWhatShapingRT2016].

@@

We used `r my_r_citation$r` for all reported analyses. Web links are printed in bold. The content of this tutorial is mainly based on @allisonSurvivalAnalysisUsing2010, @singerAppliedLongitudinalData2003, @mcelreathStatisticalRethinkingBayesian2018, @kurzAppliedLongitudinalDataAnalysis2023, and @kurzStatisticalRethinkingSecondEd2023.

@@

  These results show that this participant is initially responding to the prime even though (s)he was instructed to only respond to the target, that response competition emerges in the incongruent prime condition around 300 ms, and that only later response are fully controlled by the target stimulus. Qualitatively similar results were obtained for the other five participants. Also, in their second Experiment, @panisWhatShapingRT2016 showed that the negative compatibility effect in the mask-present conditions is time-locked to mask onset. This example shows that a simple difference between two means fails to reveal the dynamic behavior people display in many experimental paradigms [@panisHowCanWe2020; @panisNeuropsychologicalEvidenceTemporal2017; @panisStudyingDynamicsVisual2020; @panisTimecourseContingenciesPerceptual2009; @panisWhenDoesInhibition2022; @schmidtResponseInhibitionNegative2022]. In other words, statistically controlling for the passage of time during data analysis is equally important as experimental control during the design of an experiment, to better understand human behavior in experimental paradigms. As we will show in Tutorials 2 and 3, statistical models for h(t) and ca(t) can each be implemented as generalized linear mixed regression models predicting event occurrence (1/0) and response accuracy (1/0) in each bin of a selected time range, respectively.

>>>>

---
title             : "A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data"
shorttitle        : "Tutorial on hazard analysis"
author: 
  - name          : "Sven Panis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "ETH GLC, room G16.2, Gloriastrasse 37/39, 8006 Zürich"
    email         : "sven.panis@hest.ethz.ch"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Richard Ramsey"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "ETH Zürich"
authornote: |
  Neural Control of Movement lab, Department of Health Sciences and Technology (D-HEST).
  <!-- Enter author note here. -->
abstract: |
  Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. The orthodox method for analysing such data - comparing means with analysis-of-variance - is actually hiding a lot of information about psychological effects, such as their onset time and duration, and whether they are time-locked to stimuli. Such information can change key conclusions about psychological processes and can be revealed by using distributional measures.  
  Here we provide a set of tutorials on how to implement one particular distributional method known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. We illustrate how one can calculate the descriptive statistics, and how one can implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. The R code is publicly available on Github and OSF, and can easily be adapted for other data sets. We further discuss how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time hazard analysis.

@@

```{r setup, include = FALSE}
# Tutorials 1 and 4
pkg1 <- c("papaja", "citr", "tidyverse", "RColorBrewer", "patchwork")

@@

# Tutorial 3
pkg3 <- c("lme4", "nlme")

lapply(pkg3, library, character.only = TRUE)

@@

## Means versus distributional shapes

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. However, differences in means conceal when an experimental effect starts, how long it lasts, how it evolves over increasing waiting time, and whether its onset is time-locked to other events. Such information is useful not only for interpretation, but also for cognitive psychophysiology and computational model selection [@panisAnalyzingResponseTimes2020].
As a simple illustration, Figure 1 shows three examples of how an observed difference in mean response times (RTs) between two experimental conditions conceals differences in the shapes of the underlying RT distributions.

(ref:descr-fig1-caption) Means versus distributional shapes.

```{r plot1, fig.cap = "(ref:descr-fig1-caption)", out.width="80%"}
knitr::include_graphics("../Figures/rt_haz_combi_2.jpeg")
```

In each example, the mean RT is lower in condition 2 compared to condition 1. However, the distributions in the first example show that the effect starts around 200 ms and is gone by 600 ms. In the second example, the distributional effect starts around 400 ms and is gone by 800 ms. And in the third example, the distributional effect reverses around 550 ms.

## Outline of the paper

In this paper we focus on a distributional method known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis.
We first provide a brief overview of hazard analysis to orient the reader to the basic concepts and ideas that we will use throughout the paper. However, this will remain relatively short, and for detailed treatment, see @singerAppliedLongitudinalData2003, @allisonDiscreteTimeMethodsAnalysis1982, and @allisonSurvivalAnalysisUsing2010.

We then provide four different tutorials, each of which is written in R code and publicly available on Github and the Open Science Framework (OSF). The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply the analyses to their own time-to-event data sets. In Tutorial 1 we illustrate how to calculate the descriptive statistics for a published data set when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 2 we illustrate how one can fit Bayesian hazard models to the data. After selecting the best of four models, we plot the model-based effects and the model fits for a few subjects. In Tutorial 3 we illustrate how to fit hazard models in a frequentist framework. We compare the model-based effects between Bayesian and frequentist approaches. In Tutorial 4 we illustrate how to calculate the descriptive statistics when there are two independent variables.

# Overview of hazard analysis

To apply event history analysis (EHA), one must be able to define the event of interest (any qualitative change that can be situated in time, e.g., a button press, a saccade onset, a fixation offset, etc.), time point zero (e.g., target stimulus onset, fixation onset), and measure the passage of time between time point zero and event occurrence in discrete or continuous time units. Both the definition of hazard and the type of models employed depend on whether one is using continuous or discrete time units.

@@

Similarly, after dividing time in discrete, contiguous time bins indexed by t, let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular experimental condition. Discrete-time EHA focuses on the discrete-time hazard function h(t) = P(RT = t| RT $\geq$ t) and the discrete-time survivor function S(t) = P(RT $>$ t) = [1-h(t)].[1-h(t-1)].[1-h(t-2)]...[1-h(1)], and not on the probability mass function p(t) = h(t).S(t-1) and the cumulative distribution function F(t) = 1-S(t). The discrete-time hazard probability function gives you the probability that the event occurs (sometime) in bin t, given that the event has not occurred yet in previous bins. Unlike the discrete-time hazard function, which assesses the unique risk associated with each time bin, the discrete-time survivor function cumulates the bin-by-bin risks of event *non*occurrence.

For two-choice RT data, the discrete-time hazard function can be extended with the discrete-time conditional accuracy function ca(t) = P(correct | RT = t), which gives you the probability that a response is correct given that it has been emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. This latter function is also known as the micro-level speed-accuracy tradeoff function.

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons. First, as discussed by @holdenDispersionResponseTimes2009, “probability density functions can appear nearly identical, both statistically and to the naked eye, and yet are clearly different on the basis of their hazard functions (but not vice versa). Hazard functions are thus more diagnostic than density functions” (p. 331).
Second, because RT distributions may differ from one another in multiple ways, @townsendTruthConsequencesOrdinal1990 developed a dominance hierarchy of statistical differences between two arbitrary distributions A and B. For example, if F~A~(t) > F~B~(t) for all t, then both cumulative distribution functions are said to show a complete ordering. Townsend (1990) showed that a complete ordering on the hazard functions —$\lambda$~A~(t) > $\lambda$~B~(t) for all t— implies a complete ordering on both the cumulative distribution and survivor functions —F~A~(t) > F~B~(t) and S~A~(t) < S~B~(t)— which in turn implies an ordering on the mean latencies —mean A < mean B. In contrast, an ordering on two means does not imply a complete ordering on the corresponding F(t) and S(t) functions, and a complete ordering on these latter functions does not imply a complete ordering on the corresponding hazard functions. This means that stronger conclusions can be drawn from data when comparing the hazard functions using EHA. For example, when mean A < mean B, the hazard functions might show a complete ordering (i.e., for all t), a partial ordering (e.g., only for t > 300 ms, or only for t < 500 ms), or they may cross each other one or more times.
Third, EHA does not discard right-censored observations when estimating hazard functions, that is, trials for which we do not observe a response during the data collection period so that we only know that the RT must be larger than some value. This is important because although a few right-censored observations are inevitable in most RT tasks, a lot of right-censored observations are expected in experiments on masking, the attentional blink, and so forth.
Fourth, hazard modeling allows incorporating time-varying explanatory covariates such as heart rate, electroencephalogram (EEG) signal amplitude, gaze location, etc. (Allison, 2010) which is useful for cognitive psychophysiology [@meyerModernMentalChronometry1988].
Finally, as explained by @kelsoOutlineGeneralTheory2013, it is crucial to first have a precise description of the macroscopic behavior of a system (here: h(t) and ca(t) functions) in order to know what to derive on the microscopic level. For example, fitting parametric functions or computational models to data without studying the shape of the discrete-time h(t) and ca(t) functions can miss important features in the data [@panisStudyingDynamicsVisual2020; @panisWhatShapingRT2016].

@@

We used `r my_r_citation$r` for all reported analyses. Web links are printed in bold. The content of the tutorials is mainly based on @allisonSurvivalAnalysisUsing2010, @singerAppliedLongitudinalData2003, @mcelreathStatisticalRethinkingBayesian2018, @kurzAppliedLongitudinalDataAnalysis2023, and @kurzStatisticalRethinkingSecondEd2023.

@@

  These results show that this participant is initially responding to the prime even though (s)he was instructed to only respond to the target, that response competition emerges in the incongruent prime condition around 300 ms, and that only later response are fully controlled by the target stimulus. Qualitatively similar results were obtained for the other five participants. Also, in their second Experiment, @panisWhatShapingRT2016 showed that the negative compatibility effect in the mask-present conditions is time-locked to mask onset. This example shows that a simple difference between two means fails to reveal the dynamic behavior people display in many experimental paradigms [@panisHowCanWe2020; @panisNeuropsychologicalEvidenceTemporal2017; @panisStudyingDynamicsVisual2020; @panisTimecourseContingenciesPerceptual2009; @panisWhenDoesInhibition2022; @schmidtResponseInhibitionNegative2022]. In other words, statistically controlling for the passage of time during data analysis is equally important as experimental control during the design of an experiment, to better understand human behavior in experimental paradigms. As we will show in Tutorials 2 and 3, statistical models for h(t) can each be implemented as generalized linear mixed regression models predicting event occurrence (1/0) in each bin of a selected time range.

### 2024-09-18 manuscript/manuscript.tex
update manuscript

\abstract{%
Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. To move beyond mean performance measures, various distributional analyses have been proposed. Here we focus on one particular distributional analysis known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. Across four tutorials that we make publicly available on Github and OSF, we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. Along the way, we discuss how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time survival analysis.
}

@@

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating mean average performance across a series of trials. However, differences in means conceal when an experimental effect starts, how long it lasts, and whether its onset is time-locked to other events. Such information is useful not only for interpretation, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).
In this tutorial we focus on a distributional method for analyzing time-to-event data that is known as discrete-time event history analysis (EHA), a.k.a. survival, hazard, duration, failure-time, and transition analysis (Allison, 1982, 2010; Singer \& Willett, 2003). Across four tutorials that we make publicly available on Github and the Open Science Framework (OSF), we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4.

To apply EHA, one must be able to define the event of interest (any qualitative change that can be situated in time, e.g., a button press, saccade onset, fixation offset), time point zero (e.g., target stimulus onset, fixation onset), and measure the passage of time between time point zero and event occurrence in discrete or continuous time units.

@@

Similarly, after dividing time in discrete, contiguous time bins indexed by t, let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular experimental condition. Discrete-time EHA focuses on the discrete-time hazard function h(t) = P(RT = t\textbar{} RT \(\geq\) t) and the discrete-time survivor function S(t) = P(RT \(>\) t) = {[}1-h(t){]}.{[}1-h(t-1){]}.{[}1-h(t-2){]}\ldots{[}1-h(1){]}, and not on the probability mass function p(t) = h(t).S(t-1) and the cumulative distribution function F(t) = 1-S(t). The discrete-time hazard probability function gives you the probability that the event occurs (sometime) in bin t, given that the event has not occurred yet in previous bins. Unlike the hazard function, which assesses the unique risk associated with each time bin, the survivor function cumulates the bin-by-bin risks of event \emph{non}occurrence.
For two-choice RT data, the discrete-time hazard function can be extended with the conditional accuracy function ca(t) = P(correct \textbar{} RT = t), which gives you the probability that a response is correct given that it has been emitted in time bin t (Allison, 2010; Kantowitz \& Pachella, 2021; Wickelgren, 1977). This latter function is also known as the micro-level speed-accuracy tradeoff function.

@@

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. Web links are printed in bold. The content of this tutorial is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2018), Kurz (2023a), and Kurz (2023b).

@@

@@

>>>>

\abstract{%
Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. The orthodox method for analysing such data - comparing means with analysis-of-variance - is actually hiding a lot of information about psychological effects, such as their onset time and duration, and whether they are time-locked to stimuli. Such information can change key conclusions about psychological processes and can be revealed by using distributional measures.\\
Here we provide a set of tutorials on how to implement one particular distributional method known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. We illustrate how one can calculate the descriptive statistics, and how one can implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. The R code is publicly available on Github and OSF, and can easily be adapted for other data sets. We further discuss how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time hazard analysis.
}

@@

\subsection{Means versus distributional shapes}\label{means-versus-distributional-shapes}

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. However, differences in means conceal when an experimental effect starts, how long it lasts, how it evolves over increasing waiting time, and whether its onset is time-locked to other events. Such information is useful not only for interpretation, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).
As a simple illustration, Figure 1 shows three examples of how an observed difference in mean response times (RTs) between two experimental conditions conceals differences in the shapes of the underlying RT distributions.

{\centering \includegraphics[width=0.8\linewidth,height=0.67\textheight,]{../Figures/rt_haz_combi_2}

\caption{Means versus distributional shapes.}\label{fig:plot1}
\end{figure}

In each example, the mean RT is lower in condition 2 compared to condition 1. However, the distributions in the first example show that the effect starts around 200 ms and is gone by 600 ms. In the second example, the distributional effect starts around 400 ms and is gone by 800 ms. And in the third example, the distributional effect reverses around 550 ms.

\subsection{Outline of the paper}\label{outline-of-the-paper}

In this paper we focus on a particular distributional method known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis.
We first provide a brief overview of discrete-time hazard analysis to orient the reader to the basic concepts and ideas that we will use throughout the paper. However, this will remain relatively short, and for detailed treatment, see Singer and Willett (2003), Allison (1982), and Allison (2010).

We then provide four different tutorials, each of which is written in R code and publicly available on Github and the Open Science Framework (OSF). The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply the analyses to their own time-to-event data sets. In Tutorial 1 we illustrate how to calculate the descriptive statistics for a published data set when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 2 we illustrate how one can fit Bayesian hazard models to the data. After selecting the best of four models, we plot the model-based effects and the model fits for a few subjects. In Tutorial 3 we illustrate how to fit hazard models in a frequentist framework. We compare the model-based effects between Bayesian and frequentist approaches. In Tutorial 4 we illustrate how to calculate the descriptive statistics when there are two independent variables.

\section{Overview of hazard analysis}\label{overview-of-hazard-analysis}

To apply event history analysis (EHA), one must be able to define the event of interest (any qualitative change that can be situated in time, e.g., a button press, a saccade onset, a fixation offset, etc.), time point zero (e.g., target stimulus onset, fixation onset), and measure the passage of time between time point zero and event occurrence in discrete or continuous time units.

@@

Similarly, after dividing time in discrete, contiguous time bins indexed by t, let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular experimental condition. Discrete-time EHA focuses on the discrete-time hazard function h(t) = P(RT = t\textbar{} RT \(\geq\) t) and the discrete-time survivor function S(t) = P(RT \(>\) t) = {[}1-h(t){]}.{[}1-h(t-1){]}.{[}1-h(t-2){]}\ldots{[}1-h(1){]}, and not on the probability mass function p(t) = h(t).S(t-1) and the cumulative distribution function F(t) = 1-S(t). The discrete-time hazard probability function gives you the probability that the event occurs (sometime) in bin t, given that the event has not occurred yet in previous bins. Unlike the hazard function, which assesses the unique risk associated with each time bin, the survivor function cumulates the bin-by-bin risks of event \emph{non}occurrence.

For two-choice RT data, the discrete-time hazard function can be extended with the conditional accuracy function ca(t) = P(correct \textbar{} RT = t), which gives you the probability that a response is correct given that it has been emitted in time bin t (Allison, 2010; Kantowitz \& Pachella, 2021; Wickelgren, 1977). This latter function is also known as the micro-level speed-accuracy tradeoff function.

@@

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lme4} (Version 1.1.35.5; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.0; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.166; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. Web links are printed in bold. The content of the tutorials is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2018), Kurz (2023a), and Kurz (2023b).

@@

\bibitem[\citeproctext]{ref-R-lme4}
Bates, D., Mächler, M., Bolker, B., \& Walker, S. (2015). Fitting linear mixed-effects models using {lme4}. \emph{Journal of Statistical Software}, \emph{67}(1), 1--48. \url{https://doi.org/10.18637/jss.v067.i01}

\bibitem[\citeproctext]{ref-R-Matrix}
Bates, D., Maechler, M., \& Jagan, M. (2024). \emph{Matrix: Sparse and dense matrix classes and methods}. Retrieved from \url{https://CRAN.R-project.org/package=Matrix}

@@

\bibitem[\citeproctext]{ref-R-nlme}
Pinheiro, J. C., \& Bates, D. M. (2000). \emph{Mixed-effects models in s and s-PLUS}. New York: Springer. \url{https://doi.org/10.1007/b98882}
